Retrofitting Effect Handlers onto OCaml

### 2021-03-24 retro-concurrency.tex
Move example to section 3

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are given by \emph{handlers} of the effects.
For example,
%
\begin{lstlisting}
effect Read : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |Read|, which is parameterised with an input channel
of type |in_channel|, which when \emph{performed} returns a |string| value. A
computation can perform the |Read| effect without knowing how the |Read| effect
is implemented. This computation may be enclosed by different handlers that
handle |Read| differently. For example, |Read| may be implemented by performing
a blocking read on the input channel or performing the read asynchronously by
offloading it to an event loop such as |libuv|. Thanks to the separation of
effectful operations from their interpretations, effect handlers enable new
approaches to modular programming.

Effect handlers provide a structured form of delimited
continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
such as resumable exceptions, lightweight threads, coroutines, generators and
asynchronous I/O to be composably expressed. One of the primary motivations to
extend OCaml with effect handlers is to natively support composable
asynchronous I/O. Many programming languages, including OCaml, require
non-local changes to source code in order to support asynchronous I/O, often
leading to a dichotomy between synchronous and asynchronous code~\cite{colour}.
In the absence of native support for concurrent programming, OCaml developers
use monadic I/O libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where
asynchronous functions are represented as monadic computations. In these
libraries, while asynchronous functions can call synchronous functions
directly, the converse is not true. In particular, any function that calls an
asynchronous function will also have to be marked as asynchronous. Languages
such as GHC Haskell and Go provide lightweight threads, which avoids the
dichotomy between synchronous and asynchronous code. However, these languages
bake-in the lightweight thread implementation into the runtime system. With
effect handlers, asynchronous I/O can be implemented directly in OCaml as a
library without imposing a monadic form on the users.

@@

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. One standard way of implementing continuations
is to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
the compiler to utilise a CPS IR would be an enormous undertaking that would
affect the performance profile of existing OCaml applications due to the
increased memory allocations as the continuation closures get allocated on the
heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is meaningless,
and hence, we would lose compatibility with tools that inspect the program
stack. Hence, we choose not to use CPS translation and represent the
continuations as call stacks.

@@

We have implemented effect handlers in a multicore extension of the OCaml
programming language which we call \textbf{Multicore OCaml} to distinguish it
from stock OCaml. Multicore OCaml clearly delineates concurrency (overlapped
execution of tasks) from parallelism (simultaneous execution of tasks) with
separate mechanisms for expressing them. Sivaramakrishnan et
al.~\cite{Sivaramakrishnan20} describe the parallelism support in Multicore
OCaml enabled by \emph{domains}. The focus of this paper is the concurrency
support enabled by effect handlers.

The remainder of the paper continues with a description of the stock OCaml
program stack (\S\ref{sec:stack}). We then refine the design further by
focussing on retrofitting effect handlers into a mainstream \emph{systems}
language (\S\ref{sec:refine}), followed by the static and dynamic semantics for
Multicore OCaml effect handlers (\S\ref{sec:semantics}).  We then discuss the
compiler and the runtime system support for implementing effect handlers
(\S\ref{sec:impl}), and present an extensive performance evaluation of effect
handlers (\S\ref{sec:eval}) against our design goals (\S\ref{sec:req}).
Finally, we discuss the related work (\S\ref{sec:related}) and conclude
(\S\ref{sec:conc}).

@@

\section{Refining the design}
\label{sec:refine}

In order to support forking and yielding threads, we declare the
following effects:

@@

The function |spawn| (line 9) evaluates the computation |f| in an effect
handler. The computation |f| may return normally with a value, or perform
effects |Fork f'| and |Yield|. The pattern |effect Yield k| handles the effect
|Yield| and binds |k| to the continuation of the corresponding |perform|
delimited by this handler. The scheduler queue |runq| maintains a queue of
these continuations. |suspend| pushes continuations into the queue, |run_next|
pops continuations from the queue and resumes them with |()| value using the
|continue| primitive. In the case of |Yield| effect, we suspend the current
continuation |k| and resume the next available continuation. In the case of
|Fork f'| effect, we suspend the current continuation and recursively call
|spawn| on |f'| in order to run |f'| concurrently.

Effect handlers provide modularity. Observe that we can change the scheduling
algorithm from FIFO to LIFO by changing the scheduler queue to a stack. We can
implement support for synchronous read from channels by adding the following
case to the effect handler in |spawn|:

\begin{lstlisting}
let rec spawn f =
  match f () with
	...
  | effect (Read ic) k -> continue k (input_line ic)
\end{lstlisting}
%
This uses OCaml standard library's |input_line| function to read a line from
the channel |ic| and resume the continuation |k| with the resultant string.
However, performing reads synchronously blocks the entire scheduler, preventing
other threads from running until the I/O is completed.

@@

\begin{lstlisting}[numbers=left]
let run main =
  let runq = Queue.create () in
  let suspend k = Queue.push (continue k) runq in
  let pending_reads = ref [] in
  let rec run_next () =
    match Queue.pop runq with
    | f -> f ()
    | exception Queue.Empty ->
      match !pending_reads with
      | [] -> () (* no pending reads *)
      | todo ->
        let compl,todo = do_reads todo in
        List.iter (fun (str,k) ->
          Queue.push (fun () -> continue k str) runq) compl;
        pending_reads := todo;
        run_next ()
  in
  let rec spawn f =
    match f () with
    | () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
    | effect (Read ic) k ->
        pending_reads := (ic,k) :: !pending_reads;
        run_next ()
  in
  spawn main
\end{lstlisting}

The scheduler queue |runq| now holds thunks instead of continuations. The value
|pending_reads| maintains a list of pending reads and the associated
continuations (line 4). We handle the |Read| effect by pushing the pair of
input channel |ic| and continuation |k| to |pending_reads| (line 24), allowing
other threads in the scheduler to run.

@@

In this section, we refine the effect handler design for the OCaml use case. The
primary motivation for adding effect handlers to OCaml is to support a non-local
control flow mechanism to implement asynchronous I/O in direct style, generators
and enable task parallelism. These use cases do not resume the continuations
more than once. Hence, our continuations are one-shot, and resuming the
continuation more than once raises an |Invalid_argument| exception. It is
well-known that one-shot continuations can be implemented
efficiently~\cite{Bruggeman96}.

@@

@@

We can then use a top-level I/O handler, similar to the introductory example
(\S\ref{sec:aio}), to discharge the I/O operations asynchronously and resume
with the result. This handles value return cases, but what about the
exceptional cases |End_of_file| and |Sys_error|? To this end, we introduce a
|discontinue| primitive to resume a continuation by raising an exception. In
this example, on reaching the end of file, we would discontinue the captured
continuation of the |input_line| function with |discontinue k End_of_file|,
which raises the exception at |input_line|, and the open channels will be
closed.

>>>>

Effect handlers~\cite{Plotkin09} provide a modular foundation for user-defined
effects. The key idea is to separate the definition of the effectful operations
from their interpretations, which are given by \emph{handlers} of the effects.
For example,
%
\begin{lstlisting}
effect In_line : in_channel -> string
\end{lstlisting}
%
declares an \emph{effect} |In_line|, which is parameterised with an input
channel of type |in_channel|, which when \emph{performed} returns a |string|
value. A computation can perform the |In_line| effect without knowing how the
|In_line| effect is implemented. This computation may be enclosed by different
handlers that handle |In_line| differently. For example, |In_line| may be
implemented by performing a blocking read on the input channel or performing
the read asynchronously by offloading it to an event loop such as |libuv|.
Thanks to the separation of effectful operations from their interpretations,
effect handlers enable new approaches to modular programming.

Effect handlers expose a structured form of delimited
continuations~\cite{Danvy90} that enables non-local control-flow mechanisms
such as resumable exceptions, lightweight threads, coroutines, generators and
asynchronous I/O to be composably expressed. One of the primary motivations to
extend OCaml with effect handlers is to natively support composable
asynchronous I/O to enable highly scalable concurrent applications such as web
servers. Many programming languages, including OCaml, require non-local changes
to source code in order to support asynchronous I/O, often leading to a
dichotomy between synchronous and asynchronous code~\cite{colour}. In the
absence of native support for concurrent programming, OCaml developers use
monadic I/O libraries such as Lwt~\cite{lwt} and Async~\cite{async}, where
asynchronous functions are represented as monadic computations. In these
libraries, while asynchronous functions can call synchronous functions
directly, the converse is not true. In particular, any function that calls an
asynchronous function will also have to be marked as asynchronous. As a result,
large parts of the applications using these libraries tend to be in monadic
form. Languages such as GHC Haskell and Go provide lightweight threads, which
avoids the dichotomy between synchronous and asynchronous code. However, these
languages bake-in the lightweight thread implementation into the runtime
system. With effect handlers, asynchronous I/O can be implemented directly in
OCaml as a library without imposing a monadic form on the users.

@@

OCaml currently does not support any non-local control flow mechanisms other
than exceptions. This makes it particularly challenging to implement the
delimited continuations necessary for effect handlers without sacrificing the
desirable properties of OCaml. A standard way of implementing continuations is
to use continuation-passing style (CPS) in the compiler's intermediate
representation (IR)~\cite{Leijen14}. OCaml does not use a CPS IR, and changing
the compiler to utilise a CPS IR would be an enormous undertaking that would
affect the performance profile of existing OCaml applications due to the
increased memory allocations as the continuation closures get allocated on the
heap~\cite{Farvardin20}. Moreover, with CPS, stack unwinding is not meaningful,
and hence, we would lose compatibility with tools that inspect the program
stack. Hence, we choose not to use CPS translation and represent the
continuations as call stacks.

@@

We have implemented effect handlers in a multicore extension of the OCaml
programming language which we call \textbf{Multicore OCaml} to distinguish it
from \emph{stock OCaml}. Multicore OCaml delineates concurrency (overlapped
execution of tasks) from parallelism (simultaneous execution of tasks) with
distinct mechanisms for expressing them. Sivaramakrishnan et
al.~\cite{Sivaramakrishnan20} describe the parallelism support in Multicore
OCaml enabled by \emph{domains}. The focus of this paper is the concurrency
support enabled by effect handlers.

The remainder of the paper continues with a description of the stock OCaml
program stack (\S\ref{sec:stack}). We describe effect handlers in Multicore
OCaml focussing on the challenges in retrofitting them into a mainstream
\emph{systems} language (\S\ref{sec:refine}), followed by the static and
dynamic semantics for Multicore OCaml effect handlers (\S\ref{sec:semantics}).
We then discuss the compiler and the runtime system support for implementing
effect handlers (\S\ref{sec:impl}), and present an extensive performance
evaluation of effect handlers (\S\ref{sec:eval}) against our design goals
(\S\ref{sec:req}). Finally, we discuss the related work (\S\ref{sec:related})
and conclude (\S\ref{sec:conc}).

@@

\section{Effect Handlers}
\label{sec:refine}

In this section, we describe the effect handlers in Multicore OCaml, and refine
the design to retrofit them onto OCaml.

\subsection{Asynchronous I/O}

Since our primary motivation is to enable composable asynchronous I/O, let us
implement a simple lightweight thread library with support for forking new
threads and yielding control to other threads. We will then extend this library
with support for synchronously reading from channels and subsequently make it
asynchronous \textit{without changing the client code for asynchrony}. In order
to support forking and yielding threads, we declare the following effects:

@@

The function |spawn| (line 9) evaluates the computation |f| in an effect
handler. The computation |f| may return normally with a value, or perform
effects |Fork f'| and |Yield|. The pattern |effect Yield k| handles the effect
|Yield| and binds |k| to the continuation of the corresponding |perform|
delimited by this handler. The scheduler queue |runq| maintains a queue of
these continuations. |suspend| pushes continuations into the queue, |run_next|
pops continuations from the queue and resumes them with |()| value using the
|continue| primitive. In the case of |Yield| effect, we suspend the current
continuation |k| and resume the next available continuation. In the case of
|Fork f'| effect, we suspend the current continuation and recursively call
|spawn| on |f'| in order to run |f'| concurrently. Observe that we can change
the scheduling algorithm from FIFO to LIFO by changing the scheduler queue to a
stack.

We can implement support for synchronous read from channels by adding the
following case to the effect handler in |spawn|:

\begin{lstlisting}
let rec spawn f =
  match f () with
	...
  | effect (In_line ic) k -> continue k (input_line ic)
\end{lstlisting}
%
This uses OCaml standard library's |input_line| function to read a line
synchronously from the channel |ic| and resume the continuation |k| with the
resultant string. However, performing reads synchronously blocks the entire
scheduler, preventing other threads from running until the I/O is completed.

@@

\begin{lstlisting}[numbers=left]
let run main =
  let runq = Queue.create () in
  let suspend k = Queue.push (continue k) runq in
  let pending_reads = ref [] in
  let rec run_next () =
    match Queue.pop runq with
    | f -> f ()
    | exception Queue.Empty ->
      match !pending_reads with
      | [] -> () (* no pending reads *)
      | todo ->
        let compl,todo = do_reads todo in
        List.iter (fun (str,k) ->
          Queue.push (fun () -> continue k str) runq) compl;
        pending_reads := todo;
        run_next ()
  in
  let rec spawn f =
    match f () with
    | () -> run_next () (* value case *)
    | effect Yield k -> suspend k; run_next ()
    | effect (Fork f') k -> suspend k; spawn f'
    | effect (In_line ic) k ->
        pending_reads := (ic,k) :: !pending_reads;
        run_next ()
  in
  spawn main
\end{lstlisting}

The scheduler queue |runq| now holds thunks instead of continuations. The value
|pending_reads| maintains a list of pending reads and the associated
continuations (line 4). We handle the |In_line| effect by pushing the pair of
input channel |ic| and continuation |k| to |pending_reads| (line 24), allowing
other threads in the scheduler to run.

@@

Observe that this example does not resume a continuation more than once. This
also holds true for other use cases such as generators and coroutines. Hence,
our continuations are one-shot, and resuming the continuation more than once
raises an |Invalid_argument| exception. It is well-known that one-shot
continuations can be implemented efficiently~\cite{Bruggeman96}.

@@

\subsection{Resource cleanup}

@@

We can then use the |run| function that we defined earlier, to discharge the
I/O operations asynchronously and resume with the result. While this handles
value return cases, what about the exceptional cases |End_of_file| and
|Sys_error|? To this end, we introduce a |discontinue| primitive to resume a
continuation by raising an exception. In this example, on reaching the end of
file, we would discontinue the captured continuation of the |input_line|
function with |discontinue k End_of_file|, which raises the exception at
|input_line| call site, and the open channels will be closed.
